---
title: "Whatsapp sentiment analysis project"
author: "Gabriel Scozzarro"
date: "19/12/2020"
output: pdf_document
mainfont: Arial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1. Introduction
This project aim to perform a sentiment analysis on the content of a Whatsapp chat based on words and emoji used.
The time frame of the analysis is Jan 2020 - Dic 2020 

## 2. Toolbox

Importing the libraries that the project will use.

````{r lib, warning = FALSE, error = FALSE, message = FALSE}
library(rwhatsapp)
library(lubridate)
library(tidyverse)
library(tidytext)
library(kableExtra)
library(knitr)
library(ggimage)
library(RColorBrewer)
````


## 3. Preparation and reading data

After importing the contet of the chat as dataframe, some operation were performed:
1. Delete first row with the Whatsapp privicy disclaimer
2. Change one chat user name to simplify usage in code
3. Create a new feature to assign to each messages the correct season

````{r}
mychat<- rwa_read('chat_A_G.txt')
````
This is the preview of the content inside our dataframe after the processing:
````{r echo = FALSE}
mychat<- mychat[-c(1:14),] #delete first raw with whatsapp privacy encoding disclaimer



mychat$author<- as.character(mychat$author)
mychat$author[mychat$author != "Andrea Marciano"] <- "Gabriel"
mychat$author<- as.factor(mychat$author)    


mychat<- mychat %>% 
         mutate(day = date(time))%>%
         mutate(season = case_when(day >= dmy(24092019) & day <= dmy(20122019) ~ 'Autumn 2019',
                                   day >= dmy(21122019) & day <= dmy(31032020) ~ 'Winter 2020',
                                   day >= dmy(01042020) & day <= dmy(21062020) ~ 'Spring 2020',
                                   day >= dmy(22062020) & day <= dmy(23092020) ~ 'Summer 2020',
                                   day >= dmy(24092020) & day <= dmy(15122020) ~ 'Autumn 2020'
                                   ))

mychat$season<- factor(mychat$season)

mychat %>% head(10) %>% kable('latex', digits = 10, booktabs = T) %>% kable_styling(full_width = FALSE, font_size = 11, latex_options = c("striped", 'condensed','scale_down')) %>% column_spec(3, width = "25em")
````

## 4. EDA

I start the exploration and the data analysis looking for the daily messages frequency divided by season

```{r daily freq, echo=FALSE}
#Messages per seasons 
mychat %>% group_by(season) %>%
           count(day) %>%
           ggplot(aes(day, n, fill = season)) +
           geom_bar(stat = 'identity') +
           ylab('Numbers of messages') +
           xlab('season') +
           ggtitle('Messages per Seasons') +
           theme_minimal() +
           theme(legend.position = 'bottom')
```
We can observed a rise in number of messages through the season with some spikes but with essentially an exponential trend, starting from April 2020 which was in the middle of the first lockdown.
Then I explore the frequency of messages per day of week and also the frequency of messages per hour of the day.

````{r freq week, echo = FALSE}
#Messages per day of week
mychat %>% mutate(wday_num = wday(day), wday_name = weekdays(day)) %>%
           group_by(season, wday_num, wday_name) %>%
           count() %>%
           ggplot(aes(reorder(wday_name, -wday_num), n, fill = season)) +
           geom_bar(stat = 'identity') +
           xlab('') +
           coord_flip() +
           ggtitle('Messages per day of week', 'Frequency per seasons') +
           theme_minimal() +
           theme(legend.title = element_blank(), legend.position = 'bottom')

````
We can appreciate the consistency during the summer of messages, instead during the autumn 2020 we have clearly more messages per day but with 2 noticeable spikes in monday and thursday.

````{r freq hour day, echo = FALSE}
#Message frequency by the time of day
wdays<- c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday','Sunday')
names(wdays)<- 1:7 #Messages per day hours
mychat %>% mutate(hours = hour(time), wday_num = wday(day), wday_name = weekdays(day)) %>%
           count(season, wday_num, wday_name, hours) %>%
           ggplot(aes(hours, n, fill = season)) +
           geom_bar(stat = 'identity') +
           ylab('Number of messages') +
           xlab('Hours') +
           ggtitle('Number of messages per day hours', 'Frequency per seasons') +
           facet_wrap(~wday_num, ncol = 7, labeller = labeller(wday_num = wdays)) +
           theme_minimal() +
           theme(legend.title = element_blank(), legend.position = 'bottom', 
                 panel.spacing.x = unit(0.0, 'lines'))
````
Now I can move on exploring who sent more messages during the all time frame in analysis and also divided by season

````{r more messages, echo = FALSE, warning = FALSE}
#Who has sent the most messages?
mychat %>% mutate(day = date(time)) %>%
           group_by(season) %>%
           count(author) %>%
           ggplot(aes(reorder(author,n), n, fill = season)) +
           geom_bar(stat = 'identity') +
           ylab('Total number of messages') +
           xlab('User') +
           coord_flip() +
           ggtitle('Total number of messages per user', 'Who has sent the most messages?, Freq per season') +
           theme_minimal() +
           theme(legend.title = element_blank(), legend.position = 'bottom')

````
One of the user sent sensibly more messages but this can be due to the fragmentation of the content in several more small messages.
We can investigate further to see how it is.

````{r len messages, echo = FALSE, warning = FALSE, message = FALSE}
#Lenght of messages
mychat %>% mutate(text_len = nchar(text)) %>%
           group_by(author) %>%
           summarise(avg_txt_len = mean(text_len)) %>%
           ggplot(aes(author, avg_txt_len, fill = author)) +
           geom_bar(stat = 'identity') +
           xlab('Author') +
           ylab('Average messages lenght') +
           coord_flip() +
           ggtitle('Average messages lenght by author') +
           theme_minimal() +
           theme(legend.title = element_blank(), legend.position = 'bottom')

````
As we suspected the average length of messages of the other user is longer sacrificing the quantity of messages.
This can improve also the clarity of the content.
Moving forward to the content inside the chat we can start with the non text related content which are emojis.

````{r most used emoji, echo = FALSE}
emojiplot<- mychat %>% 
            unnest(c(emoji, emoji_name)) %>%
            mutate(emoji = str_sub(emoji, end = 1)) %>%
            mutate(emoji_name = str_remove(emoji_name, ':.*')) %>%
            count(emoji, emoji_name) %>%
            top_n(30, n) %>%
            arrange(desc(n)) %>%
            mutate(emoji_url = map_chr(emoji, ~paste0('https://abs.twimg.com/emoji/v2/72x72/',
                                                      as.hexmode(utf8ToInt(.x)),'.png')))
emojiplot %>% ggplot(aes(reorder(emoji_name, n), n)) +
              geom_col(aes(fill = n), show.legend = FALSE, width = .2) +
              geom_point(aes(color = n), show.legend = FALSE, size = 3) +
              geom_image(aes(image = emoji_url), size = .045) +
              ylab('Number of times emoji was used') +
              xlab('Emoji meaning') +
              ggtitle('Most used emoji') + 
              coord_flip() +
              theme_minimal() +
              theme()
````
Those are the 30 most used emojis overall inside the chat.
In the next plot we can also appreciate the most used ones by user.

````{r most used emoji by author, echo = FALSE}
emojiplot2<- mychat %>% 
             unnest(c(emoji, emoji_name)) %>%
             mutate(emoji = str_sub(emoji, end = 1))%>%
             count(author, emoji, emoji_name, sort = TRUE) %>%
             group_by(author) %>%
             top_n(8, n) %>%
             slice(1:8) %>%
             mutate(emoji_url = map_chr(emoji, ~paste0('https://abs.twimg.com/emoji/v2/72x72/', 
                                                        as.hexmode(utf8ToInt(.x)),'.png')))
emojiplot2 %>% ggplot(aes(reorder(emoji, -n), n)) +
               geom_col(aes(fill = author, group = author), show.legend = FALSE, width = .20) +
               geom_image(aes(image = emoji_url), size = .08) +
               xlab('Emiji') +
               ylab('Number of time emoji was used') +
               facet_wrap(~author, ncol = 5, scales = 'free') +
               ggtitle('Most used emoji by user') +
               theme_minimal() +
               theme(axis.text.x = element_blank())
````
Now is time for text related content. I define a list of words which will not be taken in consideration because are pronouns or articles and ecc.

````{r most used words, echo = FALSE}
#3.6 Most used words ----
useless_words<-c('il','lo','la','un','uno','una','quello','quella','quelli','nostro','vostro','di','quanto','che','se','sono',
                 'loro','alla','alle','niente','meno','piu','qui','qua','con','voi','chi','mio','tuo','va','ma','è','stata',
                 'per', 'nn','a','le','te','in','e','sto','da','sei','me','ho','ha','mi','we','per','non','sta','o','fra',
                 'su','so','hai','ci','mo','sn','eh','ti','c3','i','fa','al','ne','del')

mychat %>% unnest_tokens(input = text, output = word) %>%
           filter(!word %in% useless_words) %>%
           count(word) %>%
           top_n(30, n) %>%
           arrange(desc(n)) %>%
           ggplot(aes(reorder(word, n), n, fill = n, color = n)) +
           geom_col(show.legend = FALSE, width = .1) +
           geom_point(show.legend = FALSE, size = 3) +
           ggtitle('Most used words in chat') +
           xlab('Words') +
           ylab('Number of time it was used') +
           coord_flip() +
           theme_minimal()
````
Overall surprisingly the most used word taking in account both user is "yes".
I further divided the most used words by the user.

````{r most used words by user, echo = FALSE}
#Most used words in chat, by user
mychat %>% unnest_tokens(input = text, output = word) %>%
           filter(!word %in% useless_words) %>%
           count(author, word, sort = TRUE) %>%
           group_by(author) %>%
           top_n(20, n) %>%
           slice(1:20) %>%
           ungroup() %>%
           arrange(author, desc(n)) %>%
           mutate(order = row_number()) %>%
           ggplot(aes(reorder(word, n), n, fill = author, color = author)) +
           geom_col(show.legend = FALSE, width = .1) +
           geom_point(show.legend = FALSE, size = 3) +
           xlab('Words') +
           ylab('Number of time it was used') +
           coord_flip() +
           facet_wrap(~author, ncol = 3, scales = 'free') +
           ggtitle('Most used words by user') +
           theme_minimal()
````
Surprisingly for one of the user the 2 most user words are synonyms, meaning "yes".

## 5. Lexicon analysis
Let's see the user lexicon diversity to understand if the counting of words is due to vocabulary poorness.

````{r lexicon div, echo = FALSE}
#1.1 Lexicon diversity ----
mychat %>% unnest_tokens(input = text, output = word) %>%
           filter(!word %in% useless_words)  %>%
           group_by(author) %>%
           summarise(lex_div = n_distinct(word)) %>%
           arrange(desc(lex_div)) %>%
           ggplot(aes(reorder(author, lex_div), lex_div, fill = author)) +
           geom_col(show.legend = FALSE) +
           geom_text(aes(label = scales::comma(lex_div)), hjust = -0.1) +
           ylab("Lexicon diversity") +
           xlab("User") +
           ggtitle("Lexicon diversity in chat by author") +
           coord_flip()
````
As we see one user has a more rich vocabulary to choose on and so a few chances to repeat the same word. 

## 6. Sentiment analysis
Now is time to investigate the sentiment analysis. I started with the sentiment express in the non textual content which are emojis.
For this analysis I used the AFINN method.

````{r emoji sentiment, echo = FALSE, warning = FALSE, message = FALSE}
# GET POSITIVE / NEGATIVE FROM LEXICON PACKAGE
lexicon_negpos<- get_sentiments("afinn")
emoji_sent_score<- mychat %>% select(emoji, emoji_name) %>% 
                              unnest(emoji, emoji_name) %>%
                              mutate(emoji = str_sub(emoji, end = 1)) %>%
                              mutate(emoji_name = str_remove(emoji_name, ":.*")) %>%
                              distinct() %>%
                              unnest_tokens(input = emoji_name, output = emoji_words) %>%
                              inner_join(lexicon_negpos, by = c("emoji_words" = "word"))

bind_cols(slice(emoji_sent_score, 01:05),
          slice(emoji_sent_score, 06:10),
          slice(emoji_sent_score, 11:15),
          slice(emoji_sent_score, 16:20),
          slice(emoji_sent_score, 21:25)) %>%
          kable(format = 'latex', digits = 5, booktabs = T) %>%
          kable_styling(full_width = FALSE, font_size = 11, latex_options = c("striped",'condensed','scale_down'))


emoji_chat<- mychat %>% unnest(emoji, emoji_name) %>% mutate( emoji = str_sub(emoji, end = 1)) %>%
                        mutate( emoji_name = str_remove(emoji_name, ":.*"))

emoji_chat<- emoji_chat %>% select(author, emoji_name) %>% unnest_tokens(input = emoji_name, output = emoji_words)

user_summary<- emoji_chat %>% inner_join(lexicon_negpos, by = c("emoji_words"="word")) %>%
                              count(author, value) %>%
                              group_by(author) %>%
                              mutate(mean = n/sum(n)) %>%
                              ungroup()

reordenar_niveles <- c(-3,-2,-1,3,2,1)
colores <- c("#d7191c","#fdae61","#ffffbf","#1a9641","#a6d96a","#ffffbf")
mis_colores <- brewer.pal(5,"RdYlGn")[c(1,2,3,5,4,3)]

user_summary %>% mutate(mean = ifelse(value<0, -mean, mean)) %>%
                 group_by(author) %>%
                 mutate(balance = sum(mean)) %>%
                 ungroup() %>%
                 mutate(value = factor(value, levels = reordenar_niveles, ordered = TRUE)) %>%
                 ggplot(aes(reorder(author, balance), mean, fill = value)) +
                 geom_bar(stat = "identity", position = "stack", show.legend = FALSE, width = .5) +
                 scale_fill_manual(values = mis_colores) +
                 xlab("User") +
                 ylab("Scale of positive and negative") +
                 coord_flip() +
                 ggtitle("Emoji sentiment analysis with AFINN") +
                 theme_minimal()
````
You can see previously a preview table with the meaning of the emoji and the corresponding polarity value, then a plot with the overall sentiment analysis in the chat.
In the next plot you will see the most frequent emotion express in the chat.

````{r most freq emotion, echo = FALSE, warning = FALSE, message = FALSE}
#Most frequent emotion
lexicon_sentiment<- get_sentiments("nrc")

emoji_emo<- mychat %>% select(emoji, emoji_name) %>%
                       unnest(emoji, emoji_name) %>%
                       mutate(emoji = str_sub(emoji, end = 1)) %>%
                       mutate(amoji_name = str_remove(emoji_name, ":.*")) %>%
                       unnest_tokens(input = emoji_name, output = emoji_words) %>%
                       inner_join(lexicon_sentiment, by = c("emoji_words" = "word")) %>%
                       filter(!sentiment %in% c("negative", "positive")) %>%
                       count(emoji, emoji_words, sentiment) %>%
                       group_by(sentiment) %>%
                       top_n(4, n) %>%
                       slice(1:4) %>%
                       ungroup() %>%
                       select(-n)

chat_sentiment<- emoji_chat %>% inner_join(lexicon_sentiment, by = c("emoji_words" = "word")) %>%
                                filter(!sentiment %in% c("negative", "positive"))

chat_sentiment %>% count(sentiment) %>%
                   ggplot(aes(reorder(sentiment, n), n)) +
                   geom_col(aes(fill = n), show.legend = FALSE, width = .1) +
                   geom_point(aes(color = n), show.legend = FALSE, size = 3) +
                   coord_flip() +
                   ylab("Number of times") +
                   xlab("Emotion") +
                   scale_fill_gradient(low="#2b83ba",high="#d7191c") +
                   scale_color_gradient(low="#2b83ba",high="#d7191c") +
                   ggtitle("Most frequent emotions expressed in chat through") +
                   theme_minimal()

````
You can see as the emoji used by one of the user are practically monotone expressing a mild negative sentiment.
The other user used mostly mild negative emojis too but with more shades. 